### 2020-10-01 latex/vlp.tex
Added KL and restated proof of VLP theorem

Because \(l_\frac{\pi}{2}(\theta)\) is smooth and it approximates
\(l^\star(\theta)\), we set our VNHC to be
\[
   h(q,p) = l - l_\frac{\pi}{2}\left(\theta(q,p)\right)
\]
We still need to prove this VNHC will inject energy into a VLP. 
We will do so by showing the dynamics (\ref{eqn:vlp-hamiltonian}) 
diverge from the origin via a so-called anti-Lyapunov function.

\begin{defn}\label{defn:anti-lyapunov}
   Let \(\dot{x} = f(x) \in R^n\) be an ODE with \(f(0) = 0\). 
   A positive definite function \(V(x)\) is \textbf{anti-Lyapunov} if 
   \[
      \dot{V}(x) = dV(x) f(x) \geq 0
   \]
   where the inequality is strict for almost every \(x \in \R^n\).
\end{defn}
\textbf{Manfredi: Does this notion already exist elsewhere? I couldn't find any
sources on it, but it seems like it should've been done before.}

By applying Lyapunov stability theory \cite{lyapunov},
we see that an anti-Lyapunov function
\(V\) is simply a Lyapunov function for the negative-time system 
\(\dot{x} = -f(x)\). In other words, \(V\) proves that the orbit \(x(-t)\) 
of the system travelling backwards in time is asymptotically converging to the
origin; hence, the system must be diverging from the origin in forward time.

From Theorem \ref{thm:zero-dynamics}, the dynamics on the constraint manifold of
the VNHC are given by \((\dot{q},\dot{p})\) with 
\(l = l_\frac{\pi}{2}(\theta)\).
Supposing the initial condition is not an equilibrium, the existence of an
anti-Lyapunov function would indicate that the curve traced out on the
\((q,p)\)-plane is diverging from the origin.
This implies that the momentum \(p\) is increasing in magnitude whenever the
curve hits the \(p\)-axis, which in turn means the VLP is gaining energy on
average.

Proving the existence of an anti-Lyapunov function for the VLP
requires the following lemma.

@@

\begin{thm}\label{thm:vlp-energy-stabilization}
   Take the VLP (\ref{eqn:vlp-hamiltonian}) with initial condition 
   \((q_0,p_0) \not \in \left\{(0,0),(\pm\pi,0)\right\}\).
   Define \(\theta := \arctan_2(p,q)\). 
   A VNHC of the form \(h(q,p) = l - l(\theta)\) injects energy if there exists 
   \(l_\text{avg} \in [\underline{l},\overline{l}]\) such that 
   \begin{equation}\label{eqn:vlp-energy-gain-condition}
      \left(l(\theta) - l_\text{avg}\right)sin(2\theta) \leq 0 \text{ }\forall \theta \in \mathbb{S}^1
   \end{equation}
   with the property that the inequality is strict for almost every \(\theta\).
   If the inequality is flipped, the VNHC dissipates energy. 
\end{thm}
\begin{proof}
    Choose, as a candidate anti-Lyapunov function, the energy for the
    average-length pendulum with zero-potential at the bottom of the swing:
    \[
       E_\text{avg}(q,p) := \frac{1}{2}\frac{p^2}{m l_\text{avg}^2} 
                    + m g l_\text{avg} (1-\cos(q))
    \]
    This is positive definite at \((0,0)\) and has derivative 
    \[
      \dot{E}_\text{avg} = \frac{-g\sin(q)p \left(l(\theta)^3 - l_\text{avg}^3\right)}
                 {l_\text{avg}^2l(\theta)^2}
    \]
    The sign of \(\sin(q)p\) is dependent only on the quadrant in the
    \((q,p)\)-plane, so it is equal to the sign of \(\sin(2\theta)\). 
    Furthermore, by Lemma \ref{lemma:sign-of-cube}:
    \[ 
       \sign{l(\theta)^3 - l_\text{avg}^3} = \sign{ l(\theta) - l_\text{avg}}
    \]
    Thus, the derivative of \(E_\text{avg}\) is almost always positive since
    \begin{align*}
       \label{eq:}
       \sign{\dot{E}_\text{avg}} &= \sign{-\sin(q)p \left(l(\theta)^3 - l_\text{avg}^3\right)} \\
                   &= -\sign{\sin(2\theta) \left(l(\theta) - l_\text{avg}\right)} \\
                   &\geq 0 \text{ (by assumption)}
    \end{align*}
    Therefore \(E_\text{avg}\) is an anti-Lyapunov function.
    Since the initial condition is not an equilibrium,
    the orbit of the VNHC in the \((q,p)\) plane is diverging away from the
    origin and the VLP is gaining energy on average.
    Flipping the inequality of (\ref{eqn:vlp-energy-gain-condition}) 
    means \(E_\text{avg}\) is a regular Lyapunov function, so the VLP is
    losing energy while converging to \((0,0)\).
 \end{proof}

>>>>

Because \(l_\frac{\pi}{2}(\theta)\) is smooth and it approximates
\(l^\star(\theta)\), we set our VNHC to be
\[
   h(q,p) = l - l_\frac{\pi}{2}\left(\theta(q,p)\right)
\]
We still need to prove this VNHC will inject energy into a VLP. 
Before we can do this, we require the following lemma.

@@

Supposing the initial condition is not an equilibrium, we will show that the
constrained dynamics trace out a curve on the \((q,p)\)-plane which is diverging
from the origin.  
This implies that the momentum \(p\) is increasing in magnitude whenever the
curve hits the \(p\)-axis, which in turn means the VLP is gaining energy on
average.

\begin{thm}\label{thm:vlp-energy-stabilization}
   Take the VLP (\ref{eqn:vlp-hamiltonian}) with initial condition 
   \((q_0,p_0) \not \in \left\{(0,0),(\pm\pi,0)\right\}\).
   Define \(\theta := \arctan_2(p,q)\). 
   A VNHC of the form \(h(q,p) = l - l(\theta)\) injects energy if there exists 
   \(l_\text{avg} \in [\underline{l},\overline{l}]\) such that 
   \begin{equation}\label{eqn:vlp-energy-gain-condition}
      \left(l(\theta) - l_\text{avg}\right)sin(2\theta) \leq 0 \text{ }\forall \theta \in \mathbb{S}^1
   \end{equation}
   with the property that the inequality is strict except at the
   coordinate axes.
   If the inequality is flipped, the VNHC dissipates energy. 
\end{thm}
\begin{proof}
   Let us examine the negative-time system
   \begin{equation}\label{eqn:negative-time-vlp}
      f(q,p) = 
      \begin{bmatrix}
         \dot{q} \\ \dot{p}
      \end{bmatrix} = \begin{bmatrix}
         -\frac{p}{ml(\theta)^2} \\
         mgl(\theta)\sin(q)
      \end{bmatrix}
      ,
   \end{equation}
   with \(D = (\mathcal{Q}\times \mathcal{P}) \backslash \{(\pm \pi, 0)\}\)
   the domain of interest on which we study this system.
   Choose, as a candidate Lyapunov-like function, the energy for the
   average-length pendulum with zero-potential at the bottom of the swing:
   \[
       E_\text{avg}(q,p) := \frac{1}{2}\frac{p^2}{m l_\text{avg}^2} 
           + m g l_\text{avg} (1-\cos(q))
   \]
   This is positive definite at \((0,0)\) and has compact sublevel
   sets on \(D\). The derivative with respect to (\ref{eqn:negative-time-vlp})
   is
   \[
      \dot{E}_\text{avg} = \frac{g\sin(q)p \left(l(\theta)^3 - l_\text{avg}^3\right)}
              {l_\text{avg}^2l(\theta)^2}
      .
   \]
   The sign of \(\sin(q)p\) is dependent only on the quadrant in the
   \((q,p)\)-plane, so it is equal to the sign of \(\sin(2\theta)\). 
   Furthermore, by Lemma \ref{lemma:sign-of-cube} we have
   \[ 
       \sign{l(\theta)^3 - l_\text{avg}^3} = \sign{ l(\theta) - l_\text{avg}}
       .
   \]
   The derivative of \(E_\text{avg}\) is nonnegative since
   \begin{align*}
    \sign{\dot{E}_\text{avg}} &= \sign{\sin(q)p \left(l(\theta)^3 - l_\text{avg}^3\right)} \\
                &= \sign{\sin(2\theta) \left(l(\theta) - l_\text{avg}\right)} \\
                &\leq 0 \text{ (by assumption)}
      ,
   \end{align*}
   which means \(E_\text{avg}\) is a Lyapunov function.
   Thus, the negative-time system is stable \cite{lyapunov}; all that remains is
   to show that (\ref{eqn:negative-time-vlp}) asymptotically converges to
   \((0,0)\). 
   For this, we will use a Krazovskii-LaSalle argument
   \cite{krazovskii_lasalle}.
   First, we define 
   \(Z = \left{(q,p) \in \mathbb{S}^1\times \R \mid 
      \dot{E}_\text{avg}(q,p) = 0 \right}\).
   It is easy to show that \(Z\) is the union of the coordinate axes
   \[
      Z= \mathbb{S}^1 \times \left{0\right} \cup 
         \left{0} \times \R\right}
      .
   \]
   Then, since \(f(q,p) = (0,0)\) only when \((q,p) = (0,0)\),
   Krazovskii-LaSalle tells us that the origin is asymptotically stable on
   \(D\) for the negative-time system.
   Therefore, the orbit of the VNHC in the \((q,p)\) plane is
   diverging away from the origin in positive time, which means
   the VLP is gaining energy on average.
   Flipping the inequality of (\ref{eqn:vlp-energy-gain-condition}) 
   means \(E_\text{avg}\) is a Lyapunov function for the positive time system,
   so by the same argument the VLP is losing energy while converging
   asymptotically to \((0,0)\).
 \end{proof}
